---
title: "Introduction to R, RStudio, and Quarto"
subtitle: "A Beginner's Guide to Data Analysis"
author: "M. Yaseen"
institute: "School of Mathematical and Statistical Sciences<br>Clemson University"
format:
  revealjs:
    theme: [default, styles/custom.scss]
    logo: images/LogoSMSS.png
    footer: "Introduction to R"
    slide-number: c/t
    chalkboard: true
    touch: true
    controls: true
    width: 1600
    height: 900
    margin: 0.05
execute:
  echo: true
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| echo: false
knitr::read_chunk("R/IntroR.R")
```

# Welcome to Data Analysis!

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Three Essential Tools:**

- **R**: Free, powerful programming language for statistics
- **RStudio**: User-friendly interface for R
- **Quarto**: Modern publishing system for reports
:::

::: {.column width="48%"}
**Complete Ecosystem for:**

- Data import and cleaning
- Statistical analysis
- Professional visualizations
- Reproducible reports
- Scientific publishing
:::
:::

::: {.callout-tip}
## Why These Tools Matter
Used by millions of data scientists, researchers, and analysts worldwide for everything from academic research to business intelligence.
:::

---

## Why Use R, RStudio, and Quarto?

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Key Advantages:**

- **Completely Free** - No licensing fees ever
- **Beginner-Friendly** - Simple commands, intuitive interfaces
- **Professional Results** - Publication-ready outputs
- **Widely Used** - Industry standard tools
- **Great Community** - Extensive help and tutorials
:::

::: {.column width="48%"}
**Practical Benefits:**

- **Reproducible Research** - Share and verify analyses
- **Versatile Applications** - From simple calculations to complex modeling
- **Multiple Output Formats** - HTML, PDF, Word from one source
- **Version Control** - Track changes and collaborate effectively
:::
:::

**Bottom Line:** These tools transform how you work with data, making complex analyses accessible and professional reporting automatic.

---

# Getting Started with R

---

## What is R?

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**R is Much More Than a Calculator:**

- **Statistical Computing Environment** - Built by statisticians for data analysis
- **Data Management System** - Handle datasets of any size
- **Graphics Engine** - Beautiful, publication-ready charts
- **Programming Language** - Automate repetitive tasks
- **Statistical Toolkit** - Thousands of specialized functions
:::

::: {.column width="48%"}
**What Makes R Special:**

- **Designed for Data** - Tasks difficult elsewhere are straightforward in R
- **Extensive Libraries** - Over 19,000 packages available
- **Active Development** - Constantly updated with latest methods
- **Cross-Platform** - Works on Windows, Mac, Linux
- **Integration** - Connects with databases, web APIs, other languages
:::
:::

---

## Installing R

**Simple 5-Step Process:**

1. **Visit Official Website** - [https://cran.r-project.org/](https://cran.r-project.org/)
2. **Choose Your OS** - Windows, macOS, or Linux
3. **Download Latest Version** - Always get the most recent release
4. **Run Installer** - Use default settings for beginners
5. **Verify Installation** - Open R to confirm it works

::: {.callout-tip}
## Installation Tip
CRAN (Comprehensive R Archive Network) is the official and safest source. Avoid third-party downloads to ensure authentic, secure software.
:::

**Next Step:** While R works alone, RStudio makes everything much easier!

---

## Basic Math in R

```{r}
#| label: Ch01Code01
```

**Key Concepts:**

- **Explicit Arguments** - Use `x =` and `digits =` for clarity
- **Variable Assignment** - Use `<-` to store values
- **Function Calls** - Always include parentheses and argument names

---

## Working with Data Tables

```{r}
#| label: setup-packages
#| echo: true
#| message: false
#| warning: false
```

```{r}
#| label: Ch01Code02
```

**Why data.table?** Faster performance, intuitive syntax, memory efficient, better for beginners

---

## Basic Statistics with Fastverse

```{r}
#| label: Ch01Code03
```

**Essential Statistics Explained:**

- **Mean**: Average of all values
- **Max/Min**: Highest and lowest values  
- **Standard Deviation**: How spread out the data is
- **Count**: Number of observations

---

## Creating Professional Visualizations

```{r}
#| label: Ch01Code04
#| fig-width: 10
#| fig-height: 6
```

**Visualization Benefits:** Patterns become immediately obvious - Bob has highest grade (92), Charlie lowest (78)

---

# RStudio: Your R Interface

---

## What is RStudio?

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Integrated Development Environment (IDE):**

RStudio transforms R from a basic command line into a professional workspace with:

- **Syntax Highlighting** - Color-coded commands
- **Auto-completion** - Faster, error-free coding  
- **Project Management** - Organized workflows
- **Integrated Help** - Documentation at your fingertips
:::

::: {.column width="48%"}
**Four-Panel Layout:**

1. **Script Editor** (top-left) - Write and save code
2. **Console** (bottom-left) - Interactive R commands
3. **Environment/History** (top-right) - See variables and past commands
4. **Files/Plots/Help** (bottom-right) - Navigation and outputs
:::
:::

**Bottom Line:** RStudio makes R accessible to beginners while remaining powerful for experts

---

## Installing RStudio

**Prerequisites and Steps:**

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Before Installing:**

- **R Must Be Installed First** - RStudio requires R to function
- **Check R Installation** - Open R to verify it works

**Installation Process:**

1. Visit [https://posit.co/downloads/](https://posit.co/downloads/)
2. Choose **RStudio Desktop** (free version)
3. Download for your operating system
4. Run installer with default settings
:::

::: {.column width="48%"}
**After Installation:**

- **Launch RStudio** - You should see four-panel interface
- **Verify R Connection** - Console should show R version
- **Explore Interface** - Familiarize yourself with panels

**First Steps:**

- Create a new script (File → New File → R Script)
- Try typing `2 + 2` in console
- Use Help panel to explore documentation
:::
:::

---

## Creating and Managing Projects

```{r}
#| label: Ch01Code05
#| eval: false
```

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Why Use Projects?**

- **Organization** - Keep related files together
- **Working Directory** - Automatic folder management
- **Portability** - Easy sharing with collaborators
- **Version Control** - Git integration for tracking changes
:::

::: {.column width="48%"}
**Project Benefits:**

- **Reproducibility** - Others can run your code easily
- **Collaboration** - Share entire project folders
- **Backup** - Everything in one place
- **Scalability** - From simple analyses to complex research
:::
:::

**Best Practice:** Always work within projects - it saves time and prevents errors!

---

# Introduction to Quarto

---

## What is Quarto?

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Next Generation Publishing System:**

Quarto combines code, results, and narrative in professional documents:

- **Code Execution** - Run R code automatically
- **Dynamic Results** - Charts and tables update automatically  
- **Professional Formatting** - Publication-ready appearance
- **Multiple Formats** - HTML, PDF, Word from one source
:::

::: {.column width="48%"}
**Reproducible Research Revolution:**

**Traditional Approach Problems:**

- Analyze in one program
- Chart in another program  
- Write in word processor
- Manually copy results (error-prone!)

**Quarto Solution:**

- Everything in one document
- Automatic updates when data changes
- Complete transparency and reproducibility
:::
:::

---

## Installing Quarto

**Simple Installation Process:**

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Installation Steps:**

1. **Visit Official Site** - [https://quarto.org/docs/get-started/](https://quarto.org/docs/get-started/)
2. **Download Installer** - Choose Windows, macOS, or Linux
3. **Run with Defaults** - Standard installation handles everything
4. **Restart RStudio** - Enables Quarto integration
5. **Verify Installation** - Look for Quarto options in menus
:::

::: {.column width="48%"}
**Integration Benefits:**

- **Seamless RStudio Integration** - New file types available
- **Render Buttons** - One-click document creation
- **Preview Modes** - See results while writing
- **Project Templates** - Quick start options
- **Version Control** - Works with Git automatically
:::
:::

**Result:** Professional document creation becomes as easy as writing an email!

---

## Output Formats

::: {.columns style="gap: 1rem;"}
::: {.column width="32%"}
**HTML (Web Sharing)**
```yaml
format: html
```

**Best for:**

- Interactive sharing
- Online viewing  
- Email distribution
- Web publishing

**Features:**

- Interactive elements
- Easy sharing via links
- Mobile responsive
- Search functionality
:::

::: {.column width="32%"}
**PDF (Professional)**
```yaml
format: pdf
```

**Best for:**

- Academic papers
- Professional reports
- Print documents
- Archival purposes

**Features:**

- Page numbers
- Professional typography
- Print-ready quality
- Consistent formatting
:::

::: {.column width="32%"}
**Word (Collaboration)**
```yaml
format: docx
```

**Best for:**

- Team collaboration
- Client reviews
- Comment workflows
- Non-R users

**Features:**

- Microsoft Word compatible
- Track changes support
- Easy editing by others
- Familiar interface
:::
:::

**Power:** Write once, publish everywhere - same content, multiple professional formats!

---

## Complete Quarto Example

```{r}
#| label: Ch01Code06
#| fig-width: 10
#| fig-height: 5
```

**Key Features Demonstrated:** Automatic code execution, professional formatting, figure captioning, statistical analysis integration

---

# Working with Data Files

---

## Reading and Writing Data

```{r}
#| label: Ch01Code07
#| eval: false
```

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**File Format Comparison:**

- **CSV Files** - Plain text, widely compatible, smaller size
- **Excel Files** - Multiple sheets, formatting, larger size  
- **data.table** - R-optimized, fastest performance

**Why `fread()` over `read.csv()`?**

- Much faster performance
- Better type detection
- More flexible with delimiters
- Cleaner handling of messy data
:::

::: {.column width="48%"}
**Best Practices:**

- **Use relative paths** - "data/file.csv" not "C:/Users/..."
- **Consistent naming** - lowercase, underscores, descriptive
- **Organized folders** - separate data, scripts, outputs
- **Backup originals** - never modify raw data files
:::
:::

---

## Practical Data Analysis Example

```{r}
#| label: Ch01Code08
```

---

## Creating Comparison Visualizations

```{r}
#| label: Ch01Code09
#| fig-width: 10
#| fig-height: 6
```

**Insight:** Visualization immediately reveals that Math scores are highest on average, demonstrating the power of charts over tables alone.

---

# Essential Tips for Success

---

## Getting Help When You Need It

```{r}
#| label: Ch01Code10
#| eval: false
```

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Built-in Help System:**

- **Function Help** - `?function_name` for documentation
- **Search Help** - `help.search("topic")` for related functions
- **Examples** - `example("function")` for working code
- **Package Help** - `help(package = "packagename")` for overview
:::

::: {.column width="48%"}
**External Resources:**

- **Stack Overflow** - Huge Q&A community
- **RStudio Community** - Friendly, helpful forum  
- **R-bloggers** - Daily tutorials and tips
- **Local User Groups** - In-person networking and learning
- **Documentation Sites** - Official package guides
:::
:::

**Remember:** Every expert was once a beginner - the R community is known for being welcoming and helpful!

---

## Common Mistakes and Solutions

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Critical Mistakes to Avoid:**

1. **Case Sensitivity** - `Mean` ≠ `mean`
2. **Quotation Marks** - Text needs quotes: `"Alice"`
3. **Package Loading** - Always `library(package)` first
4. **Parentheses** - Every `(` needs a `)`
5. **Explicit Arguments** - Use `round(x = 3.14, digits = 2)`
:::

::: {.column width="48%"}
**Project Organization:**
```
my-analysis/
├── data/           # Raw data files
├── R/              # R scripts  
├── figures/        # Generated plots
├── out/            # Output files
└── README.md       # Project description
```
:::
:::

**Best Practice:** Develop good habits early - they save hours of debugging later!

---

## Keyboard Shortcuts for Efficiency

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Essential Shortcuts:**

- **Ctrl+Enter** (Win) / **Cmd+Enter** (Mac) - Run current line
- **Ctrl+Shift+Enter** - Run entire code chunk
- **Tab** - Auto-complete function names
- **Ctrl+Z** - Undo last action
- **Ctrl+Shift+C** - Comment/uncomment lines
:::

::: {.column width="48%"}
**Navigation Shortcuts:**

- **Ctrl+L** - Clear console
- **Ctrl+1** - Focus on script editor  
- **Ctrl+2** - Focus on console
- **Ctrl+S** - Save current file
- **Ctrl+Shift+N** - New script file
:::
:::

**Time Saver:** Master 3-4 shortcuts first, then gradually add more - they dramatically speed up your workflow!

---

## Troubleshooting Common Issues

```{r}
#| label: Ch01Code11
#| eval: false
```

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Package Problems:**

- **"Package not found"** - Install first: `install.packages("packagename")`
- **Loading errors** - Check package spelling and internet connection
- **Version conflicts** - Update R and packages regularly
- **Missing dependencies** - R usually installs these automatically
:::

::: {.column width="48%"}
**Data Import Issues:**

- **File not found** - Check file path and working directory
- **Encoding problems** - Try `encoding = "UTF-8"` parameter
- **Wrong delimiters** - Some "CSV" files use semicolons or tabs
- **Path problems** - Use forward slashes: `"data/file.csv"`
:::
:::

**Debug Strategy:** Read error messages carefully, Google specific errors, check documentation, ask for help - in that order!

---

# Conclusion and Next Steps

---

## What You've Accomplished Today

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Technical Skills Mastered:**

✅ **Installation** - R, RStudio, Quarto setup  
✅ **Basic Operations** - Math, statistics with fastverse  
✅ **Data Management** - Creating and manipulating data.table  
✅ **Visualizations** - Professional charts with ggplot2  
✅ **Reports** - Dynamic documents with Quarto  
✅ **Best Practices** - Explicit arguments, project organization
:::

::: {.column width="48%"}
**Conceptual Understanding:**

✅ **Reproducible Research** - Code + results + narrative  
✅ **Modern Workflow** - Projects, version control, collaboration  
✅ **Professional Output** - Multiple formats from one source  
✅ **Community Resources** - Help systems and support networks  
✅ **Troubleshooting** - Independent problem-solving skills
:::
:::

**Achievement Unlocked:** You now have the foundation for modern data science!

---

## Your Learning Journey Continues

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Immediate Next Steps:**

1. **Personal Project** - Analyze data you care about
2. **Practice Explicit Arguments** - Always use parameter names
3. **Reproduce This Tutorial** - Try with different data
4. **Experiment with Styling** - Modify colors and themes
5. **Master Basic Workflow** - Projects → Scripts → Reports
:::

::: {.column width="48%"}
**Skill Development Path:**

6. **Advanced ggplot2** - Scatter plots, histograms, faceting
7. **Data Import Mastery** - Excel, databases, web APIs
8. **Statistical Methods** - Regression, hypothesis testing
9. **Advanced Quarto** - Presentations, websites, books
10. **Package Ecosystem** - Specialized tools for your field
:::
:::

**Remember:** Every expert started exactly where you are now - the key is consistent practice!

---

## Essential Learning Resources

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Free Online Books:**

- **R for Data Science** [@wickham2016r] - The definitive beginner's guide
- **Quarto Documentation** - Comprehensive feature guide  
- **ggplot2 Book** [@wickham2016ggplot2] - Deep dive into visualization
- **fastverse Documentation** - Efficient data manipulation

**Interactive Learning:**

- **RStudio Education** - Free courses and tutorials
- **Swirl** - Learn R interactively within R
- **DataCamp** - Structured courses (some free)
:::

::: {.column width="48%"}
**Community Resources:**

- **R-bloggers** - Daily articles and tutorials
- **#RStats Twitter** - Active community sharing tips
- **Local R Meetups** - Network with other users
- **Stack Overflow** - Q&A for specific problems
- **RStudio Community** - Friendly help forum

**Professional Development:**

- **Conferences** - useR!, rstudio::conf
- **Certification** - RStudio certifications available
- **Specialized Training** - Industry-specific workshops
:::
:::

---

## Final Encouragement

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**You're Joining a Global Community:**

These tools are used daily by:

- **Data Scientists** at Google, Netflix, Facebook
- **Researchers** at universities worldwide  
- **Analysts** in government and non-profits
- **Students** in psychology, finance, biology
- **Professionals** in healthcare, marketing, sports
:::

::: {.column width="48%"}
**Remember:**

- **Everyone starts as a beginner** - You're in good company
- **The community is welcoming** - Don't hesitate to ask for help
- **Practice makes progress** - Consistent work beats perfection
- **Focus on problems you care about** - Personal interest drives learning
- **Document your journey** - Future you will thank present you
:::
:::

::: {.callout-tip}
## Your Mantra Going Forward
**Always use explicit argument names, save your work regularly, and don't hesitate to ask for help when you need it.**
:::

---

## Thank You!

::: {.columns style="gap: 2rem;"}
::: {.column width="48%"}
**Key Takeaways:**

- **Free, powerful tools** for professional data analysis
- **Reproducible research** changes how you work with data  
- **Strong community support** for continuous learning
- **Multiple output formats** from single source documents
- **Modern workflow** that scales from simple to complex
:::

::: {.column width="48%"}
**Contact & Resources:**

- **Questions?** Use RStudio Community forum
- **Advanced Help?** Stack Overflow with #r tag
- **Stay Updated?** Follow R-bloggers and #RStats
- **Local Community?** Search for R User Groups
- **Official Docs?** R, RStudio, and Quarto websites
:::
:::

**Happy Analyzing!** 🎉📊📈

---

## R Packages Used

```{r}
#| label: setup-packages1
#| ref.label: setup-packages
#| echo: true
#| message: false
#| warning: false
```

::: {.callout-note}
## Package Ecosystem Overview

**data.table** [@datatable2023] - High-performance data manipulation, much faster than base R data.frame

**fastverse** - Collection of fast, complementary packages for efficient data science workflows

**tidyverse** [@wickham2016r] - Integrated packages for data science: ggplot2, dplyr, readr, and more

**readxl** & **openxlsx** - Read and write Excel files without requiring Excel installation

**knitr** - Dynamic document generation and professional table formatting

**ggplot2** [@wickham2016ggplot2] - Grammar of graphics for beautiful, publication-ready visualizations
:::

# References

::: {#refs}
:::